---
title: "Getting started with alphaearth"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with alphaearth}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # every example needs the network and a one-time ~500 MB index download,
  # so the code is shown but not evaluated when the vignette is built.
  eval = FALSE
)
```

## Overview

[AlphaEarth Foundations](https://source.coop/tge-labs/aef) publishes annual, global **satellite embeddings**: for each 10m pixel, a 64-dimensional vector (bands `A00`-`A63`) that summarises a year of Earth Observation. The tiles are distributed as Cloud-Optimized GeoTIFFs (COGs) on [Source Cooperative](https://source.coop/tge-labs/aef).

`alphaearth` is a small, functional R Package to **find** the tiles you need and **bring them into R** as a datacube, as local GeoTIFFs, or as VRTs. The typical workflow is four steps:

## 1. Build the local index

`index()` downloads the AlphaEarth spatial index once (~500 MB) and loads it into a persistent local DuckDB database. The download and load happen a single time; later calls reuse the cache unless `overwrite = TRUE`.

```{r index}
alphaearth::index()
```

## 2. Search tiles by space and time

`search()` queries the local index and returns the matching tiles as an `sf` object (one row per tile), including the tile id, year, and both a plain HTTP `url` and a streamable `/vsicurl/` url for each COG.

```{r search}
tiles <- alphaearth::search(
  roi        = c(
    xmin = -47.9,
    ymin = -15.9, 
    xmax = -47.8, 
    ymax = -15.8
  ),
  start_date = "2020-01-01",
  end_date   = "2022-01-01"
)

tiles
```

### The `roi` argument

`roi` is flexible any of the following is accepted and reprojected to EPSG:4326 automatically:

- an `sf` / `sfc` object,
- an `sf` bounding box (`bbox`),
- a named numeric vector `c(xmin, ymin, xmax, ymax)` in EPSG:4326,
- a `terra` `SpatRaster` / `SpatVector` (its extent is used).

### The date arguments

`start_date` / `end_date` accept either a full year (`"2020"`) or an ISO date (`"2020-01-01"`). You can also select tiles directly by feature id with `fid`.

## 3. Bring the tiles into R

From here, you have three options, each covered in its own article:

### Datacubes

**`as_cube()`** build an analysis-ready data cube for [`sits`](https://e-sensing.github.io/sitsbook/) or [`stars`](https://r-spatial.github.io/stars/).

```{r tour-cube}
# sits
cube <- alphaearth::as_cube(
  x  = tiles, 
  to = "sits"
)

# or stars
cube <- alphaearth::as_cube(
  x  = tiles, 
  to = "stars"
)
```

For more information, check the `vignette("data-cubes")`

### VRTs

**`as_vrt()`** write lightweight virtual rasters (VRT) that stream from the COGs.

```{r tour-vrts}
# virtual rasters (no pixels downloaded)
vrts <- alphaearth::as_vrt(
  x = tiles,
  output_dir = "embeddings"
)
```

For more information, check the `vignette("downloading")`

### Download

**`download()`** materialise the tiles as local GeoTIFFs (optionally cropped).

```{r tour-download}
# local GeoTIFFs
files <- alphaearth::download(
  x = tiles,
  output_dir = "embeddings",
  multicores = 4
)
```

For more information, check the `vignette("downloading")`

## Data representation

Embedding values are stored as signed 8-bit integers. The de-quantization `(v / 127.5)^2 * sign(v)` is **not** applied by `alphaearth` in any step the values are returned exactly as stored, so you remain in control of any scaling.

## License & attribution

The `alphaearth` code is MIT licensed. The data is hosted by [Source Cooperative](https://source.coop/tge-labs/aef); the AlphaEarth Foundations Satellite Embedding dataset is produced by Google and Google DeepMind and licensed under CC-BY 4.0.
