---
title: "Datacubes: sits & stars"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Datacubes: sits & stars}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## From a search to a datacube

`as_cube()` turns the result of `search()` into an analysis-ready datacube for a supported backend, selected with `to`:

```{r setup}
library(alphaearth)

tiles <- alphaearth::search(
  roi    = c(
    xmin = -47.9,
    ymin = -15.9,
    xmax = -47.8,
    ymax = -15.8
  ),
  start_date = 2021,
  end_date   = 2021
)
```

Under the hood `as_cube()` never downloads pixels: for each tile it writes small **VRT** files that read the embedding COG directly over the network (via `/vsicurl/`). The heavy raster data is only touched when you actually process the cube.

Both backends share three arguments:

- `output_dir`: where the intermediate VRTs are written,
- `bands`: the embedding bands to expose (`A00`–`A63`; all 64 by default),
- `roi`: an optional crop applied to every tile (any object `search()` accepts).

## sits cubes

For [`sits`](https://e-sensing.github.io/sitsbook/), one single-band VRT is written per requested band and registered as a local `sits` cube. Extra arguments (e.g. `multicores`, `progress`) are forwarded to `sits::sits_cube()`.

```{r sits}
cube <- alphaearth::as_cube(
  x          = tiles,
  to         = "sits",
  output_dir = "cube_sits",
  bands      = c("A00", "A01", "A02"),
  multicores = 4
)

# the result is a regular sits cube, ready for the sits workflow
class(cube)
```

**Note**: In version `2.0`, [`sits`](https://e-sensing.github.io/sitsbook/) provides native access to `alphaearth`. Please, consider using their operation.

## stars objects

For [`stars`](https://r-spatial.github.io/stars/), one multi-band VRT is written per tile (the embedding dimensions become the `band` dimension) and read with `stars::read_stars()`. A single tile returns a bare `stars` object; several tiles return a named list keyed by tile id.

```{r stars}
cube <- alphaearth::as_cube(
  x          = tiles,
  to         = "stars",
  output_dir = "cube_stars",
  bands      = c("A00", "A01")
)
```

By default the cube is read **lazily** as a `stars_proxy`, meaning no pixels are read until you access them. Pass `proxy = FALSE` to pull the values into memory immediately, and any further argument is forwarded to `stars::read_stars()`.

```{r stars-eager}
# eager read (pixels loaded now)
cube <- alphaearth::as_cube(
  x      = tiles,
  to     = "stars",
  bands  = "A00",
  proxy  = FALSE
)
```

## Cropping to a region of interest

Supplying `roi` crops every tile to the same window as it is read. This is useful for when your area of interest is much smaller than a full tile:

```{r roi}
small <- alphaearth::as_cube(
  x     = tiles,
  to    = "stars",
  roi   = c(
    xmin = -47.88,
    ymin = -15.88,
    xmax = -47.85,
    ymax = -15.85
  )
)
```

## When to use what

- Reach for `as_cube()` when you want to *analyse* the embeddings in the `sits` or `stars` ecosystems without managing files yourself.

- Reach for `download()` / `as_vrt()` when you want explicit files on disk. See `vignette("downloading")`.
